Computational biology in the 21st century:\\The growth of compressive omics

### 2015-09-15 main.tex
Reworked types of biological data (again)

In the central dogma of molecular biology, DNA is transcribed into RNA, which
is translated by the ribosome into polypeptide chains, sequences of amino 
acids which singly or in complexes are known as proteins.
Proteins fold into complex, low-energy structures, which function as
cellular machines; the DNA sequence determines the amino acid sequence,
which in turn determines the folded structure of a protein.
This structure ultimately determines a protein's function within the cell.
Certain kinds of RNA also function, sometimes with proteins, as cellular 
machines.
Methods have been developed to gather biological data from every level of this process,
including sequence, abundance, structure, function, and interaction of DNA, RNA, and proteins.
It is the success of these methods that has led to the massive influx of biological data we address today.

\emph{Sequence} data, either 
nucleotide sequences (using a four-letter alphabet representing the four
DNA or RNA bases) or protein sequences (using a twenty-letter alphabet 
representing the twenty standard amino acids)
are obtained in several ways.
For protein and RNA sequence data, mass spectrometry---which
can determine protein sequence, as well as interactions---and RNA-seq---which can determine RNA sequence and abundance, 
allowing scientists to also infer the expression of the gene to which it might translate---play central roles.
However, with the advent of next-generation sequencing (NGS) technologies, the greatest volume of sequence data available are that of DNA.
For this Review, it is important to understand the structure of NGS sequence data, so we give a bit more description.

At the dawn of the genomic era, Sanger sequencing was the most widely-used 
method for reading a genome.
Though slow and labor intensive, Sanger sequencing has the advantage of being
well-tested and is still used to validate NGS results.
More recently, however, Sanger sequencing has largely been supplanted by NGS approaches,
beginning with Illumina's `sequencing by synthesis,' which allows vastly 
greater throughput due to massive parallelism, low cost, and simple sample
preparation.
Illumina sequencing and other NGS approaches
such as SOLiD, Ion Torrent, and 454 pyrosequencing do not read a single DNA
molecule end-to-end as one could read through a bound book.
Instead, DNA molecules are chopped into many small fragments;
from these fragments we generate 
\emph{reads} from one or both ends (Figure~\ref{fig:shotgun}).
These reads must be put together in the correct order 
to piece together an entire genome.
Current reads typically range from 50 to 200 bases long, though longer reads are
available with some technologies (e.g., PacBio).
Because no sequencing technology is completely infallible, sequencing machines also
provide a \emph{quality score} (or measure of the confidence in the DNA base
called) associated with each position.
Thus, an NGS read is a string of DNA letters, coupled with a string of ASCII
characters that encode the quality of the base call.
A sequencing run will produce many overlapping reads, typically stored in a FASTQ
file.

As mentioned earlier, function follows form, so in addition to
sequence and expression, structure plays an important role in biological data science.
However, we are not interested in only RNA and protein structures;
small chemical compounds represent an additional source of relevant structural data,
as they often interact with their larger RNA and protein brethren.
Determined by X-ray crystallography, NMR, electron microscopy and other 
techniques, chemical compounds---many of which are useful pharmaceuticals---can be
represented as labeled graphs (or equivalently, adjacency matrices).
In such graphs, the vertices are atoms (labeled with a chemical element) and
the edges are bonds between atoms (labeled with a bond type, such as single or
double).
Chemical compounds are stored in databases such as PubChem, and are often searched
for by medicinal chemists as potential agonists for protein targets.

When examining more complex phenomena than single proteins or compounds, we
often look to synthesize things together.
To that end, we frequently use
networks to represent biological data, such as the genetic
and physical interactions among proteins, as well as those in metabolic 
pathways~\cite{berger2013computational}.
These relationships and interactions play crucial roles in a systems-level understanding of biology.

>>>>

In the central dogma of molecular biology, DNA is transcribed into RNA, which
is translated by the ribosome into polypeptide chains, sequences of amino 
acids which singly or in complexes are known as proteins.
Proteins fold into complex, low-energy structures, which function as
cellular machines; the DNA sequence determines the amino acid sequence,
which in turn determines the folded structure of a protein.
This structure ultimately determines a protein's function within the cell.
Certain kinds of RNA also function, sometimes with proteins, as cellular 
machines.
Methods have been developed to gather biological data from every level of this process,
resulting in a massive influx of data on sequence, abundance, structure, function, and interaction of DNA, RNA, and proteins,
Much of these data are amenable to standard Big Data analysis methods; however, in this Review we focus on several examples of biological data which exhibit additional exploitable structure for creating scalable algorithms.

\emph{Sequence} data, either 
nucleotide sequences (using a four-letter alphabet representing the four
DNA or RNA bases) or protein sequences (using a twenty-letter alphabet 
representing the twenty standard amino acids)
are obtained in several ways.
For protein and RNA sequence data, mass spectrometry---which
can determine protein sequence, as well as interactions---and RNA-seq---which can determine RNA sequence and abundance, 
allowing scientists to also infer the expression of the gene to which it might translate---play central roles.
However, with the advent of next-generation sequencing (NGS) technologies, the greatest volume of sequence data available are that of DNA.
To better understand the structure of NGS sequence data, we expand a little on NGS methodologies.

At the dawn of the genomic era, Sanger sequencing was the most widely-used 
method for reading a genome.
Though slow and labor intensive, Sanger sequencing has the advantage of being
well-tested and is still used to validate NGS results.
More recently, however, Sanger sequencing has largely been supplanted by NGS approaches,
beginning with Illumina's `sequencing by synthesis,' which allows vastly 
greater throughput due to massive parallelism, low cost, and simple sample
preparation.
Illumina sequencing and other NGS approaches
such as SOLiD, Ion Torrent, and 454 pyrosequencing do not read a single DNA
molecule end-to-end as one could read through a bound book.
Instead, in so-called `shotgun sequencing' DNA molecules are chopped into many small fragments;
from these fragments we generate 
\emph{reads} from one or both ends (Figure~\ref{fig:shotgun}).
These reads must be put together in the correct order 
to piece together an entire genome.
Current reads typically range from 50 to 200 bases long, though longer reads are
available with some technologies (e.g., PacBio).
Because no sequencing technology is completely infallible, sequencing machines also
provide a \emph{quality score} (or measure of the confidence in the DNA base
called) associated with each position.
Thus, an NGS read is a string of DNA letters, coupled with a string of ASCII
characters that encode the quality of the base call.
A sequencing run will produce many overlapping reads, typically stored in a FASTQ
file.

As mentioned earlier, function follows form, so in addition to
sequence and expression, structure plays an important role in biological data science.
However, we are not interested in only RNA and protein structures;
small chemical compounds represent an additional source of relevant structural data,
as they often interact with their larger RNA and protein brethren.
Physical structures of molecules can be determined by X-ray crystallography, NMR, electron microscopy and other 
techniques.
Once determined, there are a variety of ways of representing these structures, from labeled graphs of molecular bonds to summaries of protein domains.
These representations can then be stored in databases such as PubChem or the Protein Data Bank,
and are often searched through for, e.g. potential small molecule agonists for protein targets.
Importantly, as we will expand upon later in this Review, interesting biomolecules tend to be sparse and non-randomly distributed in many representational spaces, which can be used for accelerating the aforementioned searches.

When examining more complex phenomena than single proteins or compounds, we
often look to synthesize things together.
To that end, we frequently use
networks to represent biological data, such as the genetic
and physical interactions among proteins, as well as those in metabolic 
pathways~\cite{berger2013computational}.
These relationships and interactions play crucial roles in a systems-level understanding of biology.
While standard network science tools have been employed in these analyses, metabolic pathways have additional flow constraints on their links;
as with our other examples of interesting biological data types, these additional constraints on metabolic pathways allows for more targeted approaches to gain advantages over more general techniques.
